---
title: Extracting CDC Data
output: github_document
---

This notebook documents natality data downloaded from [CDC Wonder](https://wonder.cdc.gov/natality.html){.uri} and [data.cdc.gov](https://data.cdc.gov/).

All raw, downloaded CDC data are stored in this project's **/data/** folder.

```{r setup, message = FALSE, warning = FALSE}
library(here)
library(dplyr)
library(vroom)
```

# Expanded Natality data (2016 - 2019)

Here are the latest [technical notes](https://www.cdc.gov/nchs/nvss/vsrr/natality-technical-notes.htm) on CDC's Natality data and the [CDC data dictionary](https://wonder.cdc.gov/wonder/help/Natality-expanded.html#)

## Low-risk deliveries

This dataset is inspired by the Maternal, Infant, and Child Health objective identified by [Healthy People 2030](https://health.gov/healthypeople) to *reduce cesarean births among low-risk women with no prior births* to a target of 23.6% nationwide. A low-risk birth is [defined](https://health.gov/healthypeople/objectives-and-data/browse-objectives/pregnancy-and-childbirth/reduce-cesarean-births-among-low-risk-women-no-prior-births-mich-06/data-methodology) as

-   **nulliparous**: first birth,

-   **singleton**: a single fetus (not multiple),

-   **term**: at least 37 weeks of gestation based on obstetric estimate of gestation at delivery, and

-   **vertex**: not breech / head is facing in a downward position for delivery.

We'll first get national totals for low-risk cesarean births by selecting the following criteria in CDC Wonder:

In section 10. "Select delivery characteristics"

-   **Fetal Presentation**: Cephalic

-   **Final Route and Delivery Method**: Cesarean

-   **Delivery Method Expanded**: Primary C-section; Repeat C-section; C-section (unknown if previous c-section)

In section 5. "Select pregnancy history and prenatal care characteristics"

-   **Live Birth Order**: 1

In section 12. "Select infant characteristics"

-   **OE Gestational Age Recode 11**: 37-38 weeks; 39 weeks; 40 weeks; 41 weeks; 42 weeks or more

-   **Plurality**: Single

This yields the **US Natality, 2016-2021 expanded_low-risk cesarean.txt** data file stored in **/data/**.

```{r echo=FALSE}
low_risk_cesarean_totals <- 
  vroom(
    here("data", "US Natality, 2016-2021 expanded_low-risk cesarean.txt"),
    n_max = 6,
    col_types = cols("c", "i", "i", "i"),
    delim = "\t"
  )  %>% 
  select(Year, Cesarean_births = Births) %>%  
  as.data.frame()

low_risk_cesarean_totals %>%
  mutate(Cesarean_births = prettyNum(Cesarean_births, big.mark = ",")) %>% 
  arrange(desc(Year))
```

These match the low-risk cesarean totals reported in the [Jan. 2023 National Vital Statistics Report](https://www.cdc.gov/nchs/data/nvsr/nvsr72/nvsr72-01.pdf). Next, we'll get total low-risk births so that we can calculate the rate of low-risk cesareans among these. This data is obtained by removing the filters for delivery method:

-   **Fetal Presentation**: Cephalic

-   **Live Birth Order**: 1

-   **OE Gestational Age Recode 11**: 37-38 weeks; 39 weeks; 40 weeks; 41 weeks; 42 weeks or more

-   **Plurality**: Single

```{r}
low_risk_all_deliveries_totals <- 
  vroom(
    here("data", "US Natality, 2016-2021 expanded_low-risk births.txt"),
    n_max = 6,
    col_types = cols("c", "i", "i", "i"),
    delim = "\t"
  ) %>% 
  select(Year, Births) %>%  
  as.data.frame()

low_risk_all_deliveries_totals %>%
  mutate(Births = prettyNum(Births, big.mark = ",")) %>% 
  arrange(desc(Year))
```

We can then join these into a single dataset with totals and calculate low-risk rates.

```{r}
df_low_risk_births <- 
  left_join(
    low_risk_all_deliveries_totals, 
    low_risk_cesarean_totals, 
    by = "Year"
  )

df_low_risk_births %>% 
  mutate(
    low_risk_cesarean_rate = scales::percent(Cesarean_births/Births, accuracy = .1),
    Births = prettyNum(Births, big.mark = ","),
    Cesarean_births = prettyNum(Cesarean_births, big.mark = ",")
  ) %>% 
  arrange(desc(Year))
```

For the latest rates, visit the [NCHS - VSRR Quarterly provisional estimates for selected birth indicators](https://data.cdc.gov/d/76vv-a7x8) dataset on socrata. This does not include totals but has cesarean and low-risk cesarean birth rates at the national level by race/ethnicity.
