---
title: "Extracting CDC Data"
output: github_document
---

This notebook documents our process for extracting natality data from [CDC Wonder](https://wonder.cdc.gov/natality.html){.uri} and [data.cdc.gov](https://data.cdc.gov/).

All raw, downloaded CDC data are stored in this project's **/data/** folder.

```{r setup, message = FALSE, warning = FALSE}
library(here)
library(dplyr)
library(vroom)
library(tidyr)
library(purrr)
```

# Using CDC Wonder

Expanded Natality data (2016 - 2021)

Here are the latest [technical notes](https://www.cdc.gov/nchs/nvss/vsrr/natality-technical-notes.htm) on CDC's Natality data and the [CDC data dictionary](https://wonder.cdc.gov/wonder/help/Natality-expanded.html#).

## Low-risk deliveries

This dataset is inspired by the Maternal, Infant, and Child Health objective identified by [Healthy People 2030](https://health.gov/healthypeople) to *reduce cesarean births among low-risk women with no prior births* to a target of 23.6% nationwide. A low-risk birth is [defined](https://health.gov/healthypeople/objectives-and-data/browse-objectives/pregnancy-and-childbirth/reduce-cesarean-births-among-low-risk-women-no-prior-births-mich-06/data-methodology) as

-   **nulliparous**: first birth,

-   **singleton**: a single fetus (not multiple),

-   **term**: at least 37 weeks of gestation based on obstetric estimate of gestation at delivery, and

-   **vertex**: not breech / head is facing in a downward position for delivery.

In order to get the totals for low-risk births and low-risk cesarean births we must identify the equivalent criteria in CDC Wonder's [Natality for 2016 - 2021 (expanded)](https://wonder.cdc.gov/natality-expanded-current.html) data collection. For low-risk cesarean births, we set the following:

In section 1. "Organize table layout"

-   **Group Results by**: Delivery characteristics - Year

In section 5. "Select pregnancy history and prenatal care characteristics"

-   **Live Birth Order**: 1

In section 10. "Select delivery characteristics"

-   **Year**: All Years

-   **Fetal Presentation**: Cephalic

-   **Delivery Method Expanded**: Primary C-section; Repeat C-section; C-section (unknown if previous c-section)

In section 12. "Select infant characteristics"

-   **OE Gestational Age Recode 11**: 37-38 weeks; 39 weeks; 40 weeks; 41 weeks; 42 weeks or more

-   **Plurality**: Single

Use the send button to run the query in CDC Wonder. The export of this dataset is saved as the **US Natality, 2016-2021 expanded_low-risk cesarean.txt** data file in **/data/**.

We can use vroom to load this data in R and mutate in formatted version of the counts.

```{r low-risk-cesareans}
low_risk_cesarean_totals <- 
  vroom(
    here("data", "US Natality, 2016-2021 expanded_low-risk cesarean.txt"),
    n_max = 6,
    col_types = cols("c", "i", "i", "i"),
    delim = "\t"
  )  %>% 
  select(Year, Cesarean_births = Births) %>%  
  as.data.frame()

low_risk_cesarean_totals %>%
  mutate(Cesarean_births = prettyNum(Cesarean_births, big.mark = ",")) %>% 
  arrange(desc(Year))
```

These match the low-risk cesarean totals reported on page 36, Table 17 [Jan. 2023 National Vital Statistics Report](https://www.cdc.gov/nchs/data/nvsr/nvsr72/nvsr72-01.pdf).

Next, we'll get total low-risk births so that we can calculate the proportion of low-risk cesareans among these. Note that these totals are not provided in the report linked above but they are the denominator used in the low-risk cesarean rates calculation as stated in footnote 6 of Table 17.

The low-risk birth data is obtained by using the same criteria above with the exception of changing the delivery method to All Methods:

In section 1. "Organize table layout"

-   **Group Results by**: Delivery characteristics - Year

In section 5. "Select pregnancy history and prenatal care characteristics"

-   **Live Birth Order**: 1

In section 10. "Select delivery characteristics"

-   **Year**: All Years

-   **Fetal Presentation**: Cephalic

-   **Delivery Method Expanded**: All Methods

In section 12. "Select infant characteristics"

-   **OE Gestational Age Recode 11**: 37-38 weeks; 39 weeks; 40 weeks; 41 weeks; 42 weeks or more

-   **Plurality**: Single

The export of this dataset is the **US Natality, 2016-2021 expanded_low-risk births.txt** data file stored in **/data/**. The total counts are displayed by year below.

```{r low-risk-all-deliveries}
low_risk_all_deliveries_totals <- 
  vroom(
    here("data", "US Natality, 2016-2021 expanded_low-risk births.txt"),
    n_max = 6,
    col_types = cols("c", "i", "i", "i"),
    delim = "\t"
  ) %>% 
  select(Year, Births) %>%  
  as.data.frame()

low_risk_all_deliveries_totals %>%
  mutate(Births = prettyNum(Births, big.mark = ",")) %>% 
  arrange(desc(Year))
```

We join the low-risk cesarean and low-risk births data into a single dataset and calculate the national low-risk rates from 2016 to 2021.

```{r}
df_low_risk_births <- 
  left_join(
    low_risk_all_deliveries_totals, 
    low_risk_cesarean_totals, 
    by = "Year"
  ) %>% 
  mutate(low_risk_cesarean_rate = Cesarean_births/Births)

df_low_risk_births %>% 
  mutate(
    low_risk_cesarean_rate = scales::percent(low_risk_cesarean_rate, accuracy = .1),
    Births = prettyNum(Births, big.mark = ","),
    Cesarean_births = prettyNum(Cesarean_births, big.mark = ",")
  ) %>% 
  arrange(desc(Year))
```

The total low-risk Cesarean births and low-risk Cesarean rates match the low-risk cesarean totals and percentages reported in Table 17, page 36 of the [Jan. 2023 National Vital Statistics Report](https://www.cdc.gov/nchs/data/nvsr/nvsr72/nvsr72-01.pdf).

This processed dataframe is saved as a csv in **/publish/** and RDS in **/save/** with the filename "US_low_risk_births_2016_to_2021".

```{r}
saveRDS(df_low_risk_births, here("save","US_low_risk_births_2016_to_2021.RDS"))
write.csv(df_low_risk_births, here("publish", "US_low_risk_births_2016_to_2021.csv"), row.names = FALSE)
```

For the latest rates as of 2023, visit the [NCHS - VSRR Quarterly provisional estimates for selected birth indicators](https://data.cdc.gov/d/76vv-a7x8) dataset on Socrata. This does not include totals but has cesarean and low-risk cesarean birth rates at the national level by race/ethnicity.

## State-level low-risk deliveries

In CDC Wonder we use the same criteria for low-risk cesarean births and low-risk births defined above, and add on an additional variable to Section 1. "Organize table layout" to

-   **Group Results by**: Year **And By** State of Residence

Send this result in CDC Wonder, and the output provides totals for all 50 US states and the District of Columbia from 2016 to 2021. The output files for low-risk births and low-risk cesarean delivery totals are, respectively, the **State-level Natality, 2016-2021 expanded_low-risk births.txt** and **State-level Natality, 2016-2021 expanded_low-risk cesarean.txt** data files stored in **/data/**.

These are combined into a single dataset using `vroom` and `purrr::map2_dfr` below. The low-risk cesarean rates for each jurisdiction are calculated in using a mutate.

```{r state-low-risk-totals}
state_filenames <- c(
  here("data", "State-level Natality, 2016-2021 expanded_low-risk births.txt"),
  here("data", "State-level Natality, 2016-2021 expanded_low-risk cesarean.txt")
)

datasets <- c(
  "All low-risk births",
  "Low-risk cesarean births"
)

df_low_risk_births_by_state <-
  map2_dfr(
    state_filenames, datasets,
    ~ vroom(
        file = .x,
        n_max = 306,
        col_types = cols("c", "i", "i", "c", "i", "i")
      ) %>% 
      drop_na(`State of Residence`) %>% 
      select(-Notes, -`Year Code`) %>% 
      mutate(type = .y)
  ) %>% 
  pivot_wider(
    names_from = type,
    values_from = Births
  ) %>% 
  rename(FIPS = `State of Residence Code`, State = `State of Residence`) %>% 
  mutate(low_risk_cesarean_rate = `Low-risk cesarean births`/`All low-risk births`)

df_low_risk_births_by_state
```

This processed dataframe is saved as a csv in **/publish/** and RDS in **/save/** with the filename "state_low_risk_births_2016_to_2021".

```{r}
saveRDS(df_low_risk_births_by_state, here("save","state_low_risk_births_2016_to_2021.RDS"))
write.csv(df_low_risk_births_by_state, here("publish", "state_low_risk_births_2016_to_2021.csv"), row.names = FALSE)
```

# Other data sources

CDC offers an API for US-level data from CDC Wonder:

<https://wonder.cdc.gov/wonder/help/WONDER-API.html>

We did not use this service for this project since the API is not setup to provide state-level data (or any granular geographic data). However, we did learn about the [wonderapi R package](https://github.com/socdataR/wonderapi) for querying the CDC Wonder API using R. At the time of this write-up (3/6/2023) the wonderapi package had a [branch](https://github.com/socdataR/wonderapi/tree/addD149) for adding support to query CDC's Expanded Natality dataset.

Additional sources for CDC vital statistics data include:

-   [CDC's Downloadable Files](https://www.cdc.gov/nchs/data_access/vitalstatsonline.htm#Downloadable)

-   [National Bureau of Economic Research's Public Use Data Archive](https://www.nber.org/research/data?page=1&perPage=50) - [Vital Statistics Data](https://www.nber.org/research/data?facet=datasetCat%3AVital%20Statistics&page=1&perPage=50)
